# Two Populations Means {#ch.2pops}

## Learning Objectives for Chapter

1. Compare two population means based on two samples.
1. State the null and alternative hypothesis for a two sample t-test.  
1. Determine if samples are paired or unpaired ("independent") when comparing two means.
1. Describe the three cases for comparing two populations means and determine when each one applies for a problem.
1. Perform an F-test of homogeneity of variance using samples from two populations.
1. After performing a two sample t test, sketch the t distribution, with the following parts labeled: 
  - the critical t value
  - the test statistic
  - the p value area under the curve.
  - the alpha level area under the curve. 
  - where the CI bounds would lie on the X axis (approximately)
1. Interpret the results of a t-test in terms of the original scientific question. 

## Exercises and Solutions

## Homework

## Laboratory Exercises

lab exercise

### Plant Sciences

**Submit BOTH files for your lab report using the appropriate Canvas tool**

For each part and question below, type your code in the grey area below, between the sets of back-ticks (```) to perform the desired computation and get output. Type your answers below the corresponding grey area.

### Part 1 [25 points]

Mass per mature seed (mg) of an invasive grass (medusahead, *Taeniatherum caput-medusae*) was measured when plants were grown in several randomly selected plots of perennial or annual grasses typical of the California Grasslands. The data are included in the R block below. 'A' stands for annual and 'P' for perennial.

Does the variance of seed mass differ between treatments (perennial vs. annual)? Perform a test of hypothesis at the 5% level (α=.05) using the F-statistic. Look up the critical F value in Table A.7 and by using the qf(p = 0.05, df1 = , df2 = ) function.

Complete the calculations "by hand", using only basic R functions like var(). Then, use the var.test () R function to test for difference of variances.


```{r}
seedMassA <- c(5.02, 4.34, 4.17, 7.07, 5.92, 5.33, 5.48, 4.59, 5.47, 5.88, 4.1, 5.14, 4.98, 4.47, 4.26, 5.02, 5.38, 5.3, 4.92, 4.96, 5.86, 6.65, 5.23, 4.51, 5.41, 6.23, 5.96, 5.12, 5.43, 4.98, 5.15, 5.81, 6.14, 5.87, 6.16, 5.97, 6.39, 6.25, 5.3, 5.43, 4.81, 4.76, 6.11, 4.18, 5.59, 5.26, 5.23, 5.9, 6.27, 5.31, 5.17, 4.93, 5.24, 4.96)

seedMassP <- c(4.9, 4.17, 4.47, 6.3, 4.52, 4.81, 4.4, 2.98, 4.75, 5.17, 4.64, 4.7, 5.13, 5.11, 5.33, 4.3, 4.24, 4.49, 4.46, 5.06, 4.62, 5.58, 4.39, 4.77, 5.18, 4.38, 4.76, 4.38, 4.95, 5.71, 3.03, 4.2, 4.78, 5.04, 4.76, 4.72, 4.87, 4.58, 4.69, 4.27, 5.17, 4.93, 3.51, 5.11, 5.38, 5.1, 3.2, 4.8, 4.17, 5.01, 3.95, 5.62, 5.44, 3.7, 4.08, 4.36, 4.68, 4.24)

(varA <- var(seedMassA))

(varP <- var(seedMassP))

var.test(seedMassA, seedMassP) # R function that does the complete test

(Fcalc <- varA / varP)

(r1 <- length(seedMassA))

(r2 <- length(seedMassP))

(df1 <- r1 - 1)

(df2 <- r2 - 1)

alpha <- 0.05

(Ftable <- qf(alpha, df1, df2, lower.tail = FALSE))

(p.of.Fcalc <- 2 * pf(Fcalc, df1, df2, lower.tail = FALSE)) # complete the code to get the observed significance

# The probability is multiplied by 2 because the test is two-tailed.
```

ANSWER 1.
Write the interpretation and conclusion from the test here:





### Part 2 [30 points]

Calculate the 95% confidence interval for the difference between the mean mass per seed of plants grown in annual and perennial grass plots. *Ignore the possibility of using the z-approximation due to the large sample size and use the t distribution.*

a. Based on the results of the test of equality of variances, determine what case (SG pg. 88) applies and estimate the variance of the difference between averages. Then compute the confidence interval.

b. Perform a t-test of the null hypothesis that the mass per seed does not differ between plants grown in annual or perennial grass plots. Perform all calculations "by hand" and compare to the results from using the t.test() function.


```{r}

(varAP <- (df1 * varA + df2 * varP) / (df1 + df2))

varDbar <- varAP / r1 + varAP/ r2

(tcalc <- (mean(seedMassA) - mean(seedMassP)) / sqrt(varDbar))

(ttable <- qt(alpha / 2, df = df1 + df2, lower.tail = FALSE)) # test is two-tailed

(CI.lo <- (mean(seedMassA) - mean(seedMassP)) - ttable * sqrt(varDbar))
(CI.hi <- (mean(seedMassA) - mean(seedMassP)) + ttable * sqrt(varDbar))

(t.test.1 <- t.test(seedMassA, seedMassP, alternative = "two.sided", paired = FALSE, var.equal = TRUE)) # complete code

```

ANSWER. State the extremes of the confidence interval and interpret the result of the test of hypothesis here:







### Part 3 [30 points]

Twelve plants were used in an experiment to study the effectiveness of using praying mantises to control aphid populations. Aphid density was measured before and after the addition of a mantis to the plant. Perform a test to determine if the mantis reduces aphid density. (Note that these are fictitious data and that the experimental design is simplified for teaching purposes. A real experiment should include a series of control plants to make sure that the potential change in aphid density is not due to other uncontrolled causes besides the addition of the mantis).


```{r}

aphids <- read.csv("Aphids.txt", header = TRUE)

before <- aphids$before

after <- aphids$after

(t.test.2 <- t.test(before, after, alternative = "greater", paired = TRUE)) # complete code

# repeat the test using "hand" calculations as in part 2 above. Add lines of code below.

d <- after - before

var.d <- var(d)

se.d.bar <- sqrt(var.d /   )

dfs <- 

t.calc.d <- (mean(d) - 0) / se.d.bar

ttable <- qt(alpha / 2, df =     , lower.tail = FALSE)


```

ANSWER. Interpret the result of the test of hypothesis here:



### Part 4: Paired or independent? [15 points]

For the following situations please determine if you should be conducting an independent or a paired t-test. Make sure to justify your answers!  In some of these situations multiple pieces of information are being collected so please indicate what groups/variables are being compared.

#### A
A researcher is interested in whether the presence of natural enemies significantly reduces herbivory from the invasive pest soybean aphid (Aphis glycines). She randomly chooses 36 individual soybean plants in 4 fields to measure. As a proxy for herbivory, she counts the number of soybean aphids on each plant. She then encloses half of the plants with a mesh exclusion cage, through which aphids can pass but large predatory insects cannot. After two weeks, she counts aphids on all 36 plants again, and determines the net change in aphid abundance under each treatment.

Answer here:



#### B
A master gardener wants to know whether his decision to use organic compost instead of synthetic fertilizer is going to change the yield of his broccoli. To test this, he applies a consistent amount of fertilizer or compost to the soil in each of six plots in a random design (3 receive compost, 3 receive fertilizer). He then plants the same variety of broccoli in all plots and measures the average yield (grams of broccoli head per plant) at harvest for each treatment.

Answer here:




#### C
You have developed a new cultivar of habanero chili (Capsicum chinense) that you have named "Screaming Siren". You are curious if the chilies produced at your breeding facility differ in spice from a colleague's farm in New Mexico. Both you and your colleague gather a sample of 10 random chilies and send the chilies to a lab to have their capsaicin, the compound in chilies that cause their piquancy, concentration measured. 

Answer here:




#### D
A medical research program is evaluating the efficacy of two natural ingredients, fish oil and niacin (vitamin B3), on blood triglyceride level. You wish to know if the ingredients differ in effect on blood triglycerides. Forty patients had their blood triglyceride levels measured before the experiment began. The forty patients were divided into two treatment groups of twenty patients. One treatment group received 4 grams of fish oil per day, the other received 1 gram of niacin per day.  After 30 days the patients had their blood triglyceride level measured.


Answer here:




#### E

A researcher is interested in determining the difference in plant species composition after seeding with native perennial seeds. Before seeding they sample quadrats within the field at random to determine the initial plant species composition. At the end of the second growing season they sample quadrats within the field at random to determine the new plant species composition.


### Animal Sciences

