# Analysis of Variance {#ch.anova}

## Learning Objectives for Chapter


ANOVA intuition. Present students with a series of groups of objects and ask them to guess if they think they are subsets of the some large group of whether any group seems to come from a different population.

Exercise: All objects in each of the squares below were randomly selected from a single larger set called  *parent* set. The set of objects in any square is called a *group*. Does it look like all groups have the same parent set?

```{r anova.intuition1, fig.cap="Does it look like all groups come from the same parent population?"}

# Code to generate 4 groups from two populations.

library(randomcoloR)

par(mfrow = c(2,2))

for (i in 1:4) {
  plot(rep(1:3, 3), 
         rep(1:3, each = 3),
         pch = 21,
         cex = 4,
       xlim = c(-1, 5), 
       ylim = c(0.5, 3.5),
       bg = randomColor(count = 9, 
                        hue = c(rep("orange", 3), "yellow")[i], 
                        luminosity = c(rep("light", 3), "bright")[i]), 
       axes = FALSE,
       xlab = paste("GROUP", i, sep = " "), 
       ylab = "")
  }

```

Is the answer the same for the next figure?

```{r anova.intuition2, fig.cap="Does it look like all groups come from the same parent population?"}

# Code to generate 4 groups from a single population.

par(mfrow = c(2,2))

for (i in 1:4) {
  plot(rep(1:3, 3), 
         rep(1:3, each = 3),
         pch = 21,
         cex = 4,
       xlim = c(-1, 5), 
       ylim = c(0.5, 3.5),
       bg = randomColor(count = 9, 
                        hue = "orange", 
                        luminosity = "light"), 
       axes = FALSE,
       xlab = paste("GROUP", i, sep = " "), 
       ylab = "")
  }


```


## Exercises and Solutions

## Homework

