---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Treatment Structures and Comparisons {#chTrtstr}

## Learning Objectives for Chapter {#LearnObj12}
1. Identify the treatment in a factorial experiment
1. State the null and alternative hypotheses of a factorial experiment.
1. Define and compare simple and main effects in a factorial.
1. Given a graph of factorial experiment results, determine if there are main effects and/or interaction effects. 
1. Calculate PLSD confidence intervals for a pairwise comparison of means and determine if two means are significantly different based on the CIs. 
1. Design factorial treatment combinations to answer a research question.
1. Transform a real-world question into a null hypothesis about means.
1. Create linear combinations of means that address practical real-world questions.
1. Define statistical interaction and explain it graphically.
1. Test hypotheses involving more than one and more than 2 means.
1. Interpret result of ANOVA with factorial treatments.
1. Interpret a graph of a factorial experiment, including description of L, Q and C trends.
1. Given a set of treatments, formulate and test useful contrasts.
1. Determine if a linear combination of means is a contrast.
1. Create contrasts to estimate the value and variance of interactions.
1. Estimate the variance of a linear combination of means.
1. Make confidence intervals for linear combination of estimated means.
1. Calculate and use the Least Significant Difference (LSD) to test for differences among treatments.
1. Apply the Protected LSD and explai how it ameliorates the inflation of alpha?
1. Create linear combinations of estimated treatment means.
1. Create contrasts to answer research questions such as effects of treatments, linear and quadratic effects of continuous treatment variables.


## Pairwise Comparisons {#PairComp}

## Factorials {#Factorials}

## Control vs. Treatments {#CtlTrt}

## Linear, Quadratic and Cubic Trends {LCQtrends}

## Contrasts {#Contrasts}

## Linear Combinations {#LinCombo}

## Exercises and Solutions {#ExSol12}

## Homework {#Hwk12}

## Laboratory Exercises

### Plant Sciences {#Lab12PLS}

Prepare an .Rmd document starting with the following text, where you substitute the corresponsing information for author name and date.

```
---
title: "Lab07 Factorial ANOVA and LSD"
author: "YourFirstName YourLastName"
date: "enter date here"
output: html_document
---
```

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# install.packages(c("tables", "pander","lsmeans", "multcomp","agricolae"))

library(tables)

library(pander)

library(lsmeans)

library(multcomp)

library(agricolae)

```

#### Instructions

The same instructions for completion and submission of work used in previous labs applies here. Refer to previous labs for the details. 

For this lab you have to be familiar with the following sections of the textbook:

\@ref(label) RCBD without subsamples
\@ref(label) Factorial Experiments
\@ref(label) Fisher's Protected LSD


In this exercise we analyze the data resulting from an experiment at the UC Davis Plant Sciences research fields. These are based on the **same experiment used in Lab 06**, but now we will consider that seeding and fertilizer are two **factors** that create the set of 6 treatments in a **factorial** treatment structure. We ignore the factor water for simplicity and because it did not cause any effects. The 6 treatments were applied in a **Completely Randomized Block Design** in which each of 6 treatments were applied randomly to 2 plots in each of 3 blocks for a total of 36 experimental units or plots.

Simulated block effects of 0.50, 0.25, and -0.75 were added to logMass observations in blocks 1, 2 and 3 to show block effects. Blocks were defined areas that were more or less homogeneous in soil and plant properties and that were contiguous. We will assume that the variances of the errors or residuals are the same in all treatments. Each observation represents the average mass of seeds produced by 4-5 medusahead plants (logMass). Data were log transformed as $logMass = log(SeedMass + 0.2)$ to achieve normality and equality of variances.


#### Part 1. Factorial ANOVA using R functions [ points]

##### 1a. Random assignment of treatments to plots

In a RCBD the treatments are assigned randomly to plots within blocks such that each treatment appears at least once in each block. Because we ignore the water factor, treatments will have 2 replicate plots in each block, one for each original level of the water factor. There are a total of 6 treatments that resulf from the combination each level of the factor "nitrogen" (n and N) with each of the three levels of the factor "seeding" (s, S and E). The treatments are identified by the two letters together. For example. NE is the treatment with addition of nitrogen and at the egde between naturalized annual and seeded native perennial grasses.

First, we pretend that the experiment is about to start and we need to assign treatments to experimental units. The random assignment of treatments to plots within blocks can be done using the '*'sample' R function. To preform the random assignment we start by creating a dataframe with a row for each treatment and a column for each factor. A new column called "Treatment" is created by pasting together the two letters that represent the corresponding level of the factor.

```{r}
# Read and understand the code in this part but do not type or modify anything.
# Create a data frame with all combinations of levels of the 2 factors. First, create treatment names for all 6 treatments. 

treat <- expand.grid(
  nitro = c("n", "N"), 
  seeding = c("s", "S", "E")
  )


# Create a column containing the treatment name in the data frame

treat$Treatment <- paste(
  treat$nitro, 
  treat$seeding, sep = ""
  )

```

In the second step, plots in each block are labeled with the numbers 1 through 12. We make a character vector with two copies of each treatment label, shuffle the order of treatment with 'sample' and column-bind the results with the plot labels. At this point, the objects called 'block1', 'block2' and 'block3' list the plot number and the treatment it should receive.


```{r}
# Number the 12 plots in each block 1 through 12, then shuffle the treatment names and assign to plots of each block.
# The sample() function is used to create a vector that has each treatment randomly assigned to each block twice, because there are two plots for each treatment in each block.

# Plots 1-12 of each block get:

(block1 <- cbind(1:12, sample(c(treat$Treatment, treat$Treatment))))


(block2 <- cbind(1:12, sample(c(treat$Treatment, treat$Treatment))))


(block3 <- cbind(1:12, sample(c(treat$Treatment, treat$Treatment))))

```

##### 1.b Add columns with factor levels to the data

Treatments were applied, plantts grown and measured. Data were collected and organized, and we enter them below. Because the data have the labels for the original treatments, we need to split the labels into the columns that correspond to nitrogen and seeding treatment. In order to analyze the experiment as a factorial of nitrogen x seeding, each factor has to be in a separate column.

```{r}
# Read and understand the code in this part but do not type or modify anything.
# Read data (seedF for "factorial")

# seedF <- read.csv(file = "Lab07SeedMassData.txt", header = TRUE)

seedF <- read.table(header = TRUE, text = "
Treatment  block  logMass
wnE  Block1  -0.210836005431437
wNE  Block1  1.55288938986704
WnE  Block1  0.055235139345024
WNE  Block1  0.835167804254832
wns  Block1  1.24627323200302
wnS  Block1  0.540066519204629
wNs  Block1  1.57199447282598
wNS  Block1  1.23870355136554
Wns  Block1  0.986995928669204
WnS  Block1  -0.170749893385349
WNs  Block1  0.664136397253749
WNS  Block1  0.736888706368722
wnE  Block2  0.041449411568171
wNE  Block2  1.428708818124
WnE  Block2  0.319992367391131
WNE  Block2  1.03098409320956
wns  Block2  0.643797294741899
wnS  Block2  -0.511211922892043
wNs  Block2  1.84080033644501
wNS  Block2  0.813650468511317
Wns  Block2  1.69986757761409
WnS  Block2  -0.154140967169852
WNs  Block2  1.423732115253
WNS  Block2  0.2968644855256
wnE  Block3  -0.655926955677754
wNE  Block3  -0.27934035269992
WnE  Block3  -0.5542030003591
WNE  Block3  -1.12888399571028
wns  Block3  0.45627765803375
wnS  Block3  -1.04875645623447
wNs  Block3  0.49759250163951
wNS  Block3  -0.097792008066854
Wns  Block3  -0.513584856362292
WnS  Block3  -1.36796547632004
WNs  Block3  1.02290365359422
WNS  Block3  -0.527897010482484
")

```

The two factor columns are sufficient to perform the analysis using preexisting R functions for factorial treatment combinations, but to do the calculations using only "basic" R functions (we refer to this as "by hand") we need a new column that contain labels for all 12 treatments. We create the colkumn by pasting the letters for nitrogen and seeding levels, make it into a factor, and add it to the data frame. Recall that a factor is a special type of column in data frames that behaves as a categorical variable with a certain set of possible values. The same column of strings could be set to be a character vector, in which case it would not have reconizable levels. You can explore this by running 'levels(seedF\$NStreat)' and then 'levels(as.character(seedF\$NStreat))'

```{r}
# Add columns $nitro and $seeding to the data, to show the levels for each factor.

seedF$nitro <- factor(substr(seedF$Treatment, 2, 2))

seedF$seeding <- factor(substr(seedF$Treatment, 3, 3))

# Add column NStreat for treatments resulting from nitro and seeding combinations.

seedF$NStreat <- factor(paste(seedF$nitro, seedF$seeding, sep = ""))

```


##### 1.c Perform a test of the Ho that all treatment means are equal.

For this part you need to test the null hypotheses that:
- there are no interactions between nitrogen and seeding,
- there are no "main effects" of nitro or seeding.

"Main effects" refers to the overall effect of a factor averaged over all levels of other factors. "Simple effects" are the effects of one factor for a given individual level of another factor. For example, the simple effect of S for level n is:

$$\bar{Y}_{nS}$$

We do these tests with an ANOVA that partitions all variation (sum of squares) in the response variable (logMass) into Block, Treatment and Error, and then partition the Treatment variation into main effect of nitro, main effect of seeding, and the interaction between nitro and seeding (nitro x seeding).


```{r}

# First do the ANOVA ignoring the treatment factorial structure.

# This yields the first partition of SS

rcbd.mod <- lm(formula = logMass ~ NStreat + block, data = seedF)

pander(anova(rcbd.mod))

# Now, analyze the same data, this time partitioning the Treatment SS according to a factorial treatment structure.

rcbdF.mod <- lm(formula = logMass ~ nitro * seeding + block, data = seedF)

# complete the code below to make an ANOVA table for the same data partitioning the Treatment SS according to factorial, as above:

pander(anova(rcbdF.mod))


```


Questions:

a) Report MSE in the RCBD and RCBDF models.

b) State the null hypothesis for the main effects of N in the RCBDF model.

c) State the null hypothesis for the main effects of seeding in in the RCBDF model.

c) Do we reject or fail to reject the following null hypothesis for the RCBDF model?

Ho: There is no interaction between nitrogen and seeding affecting medusahead seed mass?


#### Part 2. Factorial ANOVA detailed calculations [ points]

Create a new column in the data for each of the following:
- overall average
- nitro treatment average
- nitro effect = nitro average - overall average
- seeding treatment average
- seeding effect = seeding average - overall average
- overall average + nitro effect + seeding effect
- nitro x seeding average
- nitro x seeding interaction = 
  = nitro x seeding average - (overall average + nitro effect + seeding effect)
- block average
-block effect = block average - overall average
- residual


Then calculate the sum of squares and degrees of freedom for nitro effect, seeding effect, interaction and residuals. Make sure that the numbers match the results from the previous section.


```{r}

# First calculate the average medusahead seed mass per factor and add to a data frame

# Type the column name of the response variable after the dollar sign:

seedF$overall.avg <- mean(seedF$logMass) 

nitro.avg <- aggregate(formula = logMass ~ nitro, 
                       data = seedF, 
                       FUN = mean)

names(nitro.avg)[2] <- "nitro.avg"

seedF <- merge(seedF, nitro.avg)

# Type the function for calculating averages:

seed.avg <- aggregate(formula = logMass ~ seeding, 
                      data = seedF, 
                      FUN = mean)

names(seed.avg)[2] <- "seed.avg"

seedF <- merge(seedF, seed.avg)


nitro.seed.avg <- aggregate(formula = logMass ~ nitro + seeding, 
                      data = seedF, 
                      FUN = mean)

names(nitro.seed.avg)[3] <- "nitro.seed.avg"

seedF <- merge(seedF, nitro.seed.avg)


# Create a vector called block.avg to aggregate the response variable by block and calculate the means of each block

block.avg <- aggregate(formula = logMass ~ block, 
                       data = seedF, 
                       FUN = "mean")

names(block.avg)[2] <- "block.avg"

seedF <- merge(seedF, block.avg)


# Second, calculate the effects and residuals

seedF$nitro.fx <- seedF$nitro.avg - seedF$overall.avg

# Type the vector name for the average of the seeded/unseeded treatment:

seedF$seed.fx <- seedF$seed.avg - seedF$overall.avg 

seedF$nitro.seed.fx <- with(seedF, nitro.seed.avg - (overall.avg + nitro.fx + seed.fx))

seedF$block.fx <- seedF$block.avg - seedF$overall.avg

seedF$resdl <- with(seedF, 
                    logMass - 
                       overall.avg - 
                       nitro.fx - 
                       seed.fx - 
                       nitro.seed.fx - 
                       block.fx)


# Finally, type your equations to calculate the sum of squares.

(SSnitro <- sum(seedF$nitro.fx ^ 2))

(SSseed <- sum(seedF$seed.fx ^ 2))

(SSnitro.seed <- sum(seedF$nitro.seed.fx ^ 2))

(SSblock <- sum(seedF$block.fx ^ 2))

(SSresdl <- sum(seedF$resdl ^ 2))

# HINT: check your answers of the SS with the anova table in Part1. It should be the same if your calculations are correct:

pander(anova(rcbdF.mod)) 


```


#### Part 3. Multiple comparison of means using Fisher's PLSD [ points]

Fisher’s Protected Least Significant Difference is the LSD that is applied only if the overall F test is significant. In this section you need to calculate the Least Significant differences for nitro, seeding and nitro x seeding levels. For each one, construct a table that shows the means that are NOT significantly different connected by a common letter.

``` {r}
# First, use the LSD.test() function of the agricolae package.


LSD.test(rcbdF.mod, "nitro", group = TRUE, console = TRUE)


LSD.test(rcbdF.mod, "seeding", group = TRUE, console = TRUE)


LSD.test(rcbdF.mod, c("nitro", "seeding"), group = TRUE, console = TRUE)


# Second, calculate the LSD's and CI's "by hand."
# Hint: check your answers of LSD with the lsd test table. It should be the same if your calculations are correct.

(dfe.rcbd.F <- 36 - 1 - 2 - 2 - 2 - 1)
(t.critical <- qt(0.975, dfe.rcbd.F))
(MSE.rcbd.F <- SSresdl / dfe.rcbd.F)


# nitrogen
(se.nitro <- sqrt(MSE.rcbd.F / (length(seedF$logMass) / length(levels(seedF$nitro)))))
(LSD.nitro <- t.critical * sqrt(2) * se.nitro)


# seeding
(se.seed <- sqrt(MSE.rcbd.F / (length(seedF$logMass) / length(levels(seedF$seeding)))))

# Type your equation and calculate the LSD for seeding.
(LSD.seed <- t.critical * sqrt(2) * se.seed)


# nitrogen x seeding
(se.nitro.seed <- sqrt(MSE.rcbd.F / 
                         (length(seedF$logMass) / 
                            (length(levels(seedF$nitro)) * 
                            length(levels(seedF$seeding))))))

# Type your equation and calculate the LSD for nitro * seeding
(LSD.nitro.seed <- t.critical * sqrt(2) * se.nitro.seed)


```


Question Read the last lsd table and answer following questions:.

a) Do treatments “n:s” and “N:s” differ? 

b) Do treatments “n:s” and “N:E” differ? 

c) Do treatments “n:s” and “n:E’ differ? 





### Animal Sciences {#Lab12ANS}


